Finally, a functional class scoring method for enrichment analysis of Infinium DNA methylation data

### 2023-11-15 figs/GSE158422_sensitivity.Rmd
progress on sensitivity analysis

TESTING=TRUE

@@

#testing
SAMPLESIZES=c(22,10,3)
ssres <- lapply(SAMPLESIZES,function(n) {
  SEEDS <- seq(100,300,100)
  dsres <- mclapply(SEEDS, function(SEED) {
    set.seed(SEED)
    mysample <- sample(1:37,n,replace=FALSE)*2
    mysample <- c(mysample,(mysample-1))
    mysample <- mysample[order(mysample)]
    design2 <- design[mysample,]
    design2 <- design2[,colSums(design2)>0]
    mval2 <- mval[,mysample]
    dm2 <- runlimma(mval=mval2,design=design2,myann=myann)
    dmagg2 <- agg(dm2,cores=2)
    m2 <- dmagg2$gme_res_df[,"mean",drop=FALSE]
    lamres2 <- runmitch(m=m2,genesets=gs_symbols,cores=2)
    lamsig2 <- subset(lamres2,p.adjustANOVA<0.05)
    lamsigsets2 <- paste(rownames(lamsig2),as.character(lamsig2$s.dist > 0))
    TOT=length(lamsigsets2)
    TP=length(which(lamsigsets2 %in% lamsigsets))
    FP=length(which(!lamsigsets2 %in% lamsigsets))
    FN=length(which(!lamsigsets %in% lamsigsets2))
    RES=c("TOT"=TOT,"TP"=TP,"FP"=FP,"FN"=FN)
    RES
  },mc.cores=5)
  do.call(rbind,dsres)
})
blah()

# full
if (TESTING == FALSE) {

SAMPLESIZES=c(2,3,6,10,16,22,30)
ssres <- lapply(SAMPLESIZES,function(n) {
  SEEDS <- seq(100,300,100)
  dsres <- mclapply(SEEDS, function(SEED) {
    set.seed(SEED)
    mysample <- sample(1:37,n,replace=FALSE)*2
    mysample <- c(mysample,(mysample-1))
    mysample <- mysample[order(mysample)]
    design2 <- design[mysample,]
    design2 <- design2[,colSums(design2)>0]
    mval2 <- mval[,mysample]
    dm2 <- runlimma(mval=mval2,design=design2,myann=myann)
    dmagg2 <- agg(dm2,cores=2)
    m2 <- dmagg2$gme_res_df[,"mean",drop=FALSE]
    lamres2 <- runmitch(m=m2,genesets=gs_symbols,cores=2)
    lamsig2 <- subset(lamres2,p.adjustANOVA<0.05)
    lamsigsets2 <- paste(rownames(lamsig2),as.character(lamsig2$s.dist > 0))
    TOT=length(lamsigsets2)
    TP=length(which(lamsigsets2 %in% lamsigsets))
    FP=length(which(!lamsigsets2 %in% lamsigsets))
    FN=length(which(!lamsigsets %in% lamsigsets2))
    RES=c("TOT"=TOT,"TP"=TP,"FP"=FP,"FN"=FN)
    RES
  },mc.cores=5)
  do.call(rbind,dsres)
})

}

@@

dots <- lapply(ssres,function(x) { x[,"TP"] })
boxplot(dots,names=SAMPLESIZES,col="white",ylab="no. true positives",xlab="sample size",cex=0,main="GMEA") ; grid()
beeswarm(dots,add=TRUE,pch=1,cex=0.6)
abline(h= ssres[[1]][1,1])

@@

head(dm, 10) %>% kbl() %>% kable_paper("hover", full_width = F)

sigup <- rownames(subset(dm,logFC > 0 & adj.P.Val < 0.05))
sigdn <- rownames(subset(dm,logFC < 0 & adj.P.Val < 0.05))

@@

tic()
gsaup <- gsameth(sig.cpg=sigup, all.cpg=rownames(dm), collection=gs_entrez)
gsadn <- gsameth(sig.cpg=sigdn, all.cpg=rownames(dm), collection=gs_entrez)
toc() # 12.9 sec

@@

gsig_up <-  subset(gsaup,FDR<0.05)
gsig_up %>% kbl(caption="Hypermethylated pathways") %>% kable_paper("hover", full_width = F)
gsig_dn <-  subset(gsadn,FDR<0.05)
gsig_dn %>% kbl(caption="Hypomethylated pathways") %>% kable_paper("hover", full_width = F)
gsig <- rbind(gsig_up,gsig_dn)

@@

gsres <- lapply(SAMPLESIZES,function(n) {
  SEEDS <- seq(100,5000,100)
  dsres <- mclapply(SEEDS, function(SEED) {
    set.seed(SEED)
    mysample <- sample(1:37,n,replace=FALSE)*2
    mysample <- c(mysample,(mysample-1))
    mysample <- mysample[order(mysample)]
    design2 <- design[mysample,]
    design2 <- design2[,colSums(design2)>0]
    #mval downsample
    mval2 <- mval[,mysample]
    dm2 <- runlimma(mval2,design2,myann)
    sigup2 <- rownames(subset(dm2,adj.P.Val<0.05 & logFC>0))
    sigdn2 <- rownames(subset(dm2,adj.P.Val<0.05 & logFC<0))
    #gsamet
    gsaup2 <- gsameth(sig.cpg=sigup2, all.cpg=rownames(dm2), collection=gs_entrez)
    gsadn2 <- gsameth(sig.cpg=sigdn2, all.cpg=rownames(dm2), collection=gs_entrez)
    gsig_up2 <- subset(gsaup2,FDR<0.05)
    gsig_dn2 <- subset(gsadn2,FDR<0.05)
    gsig2 <- rbind(gsig_up2,gsig_dn2)
    TOT=nrow(gsig2)
    TP=length(which(rownames(gsig2) %in% rownames(gsig)))
    FP=length(which(!rownames(gsig2) %in% rownames(gsig)))
    TPES=0
    RES=c("TOT"=TOT,"TP"=TP,"FP"=FP,"TPES"=TPES)
    RES
  },mc.cores=8)
  do.call(rbind,dsres)
})

mx <- matrix(rep(0,200),ncol=4)
colnames(mx) <- c("TOT","TP","FP","TPES")

@@

>>>>

@@

SAMPLESIZES=c(2,3,6,10,16,22,30)
ssres <- lapply(SAMPLESIZES,function(n) {
  SEEDS <- seq(100,5000,100)
  dsres <- mclapply(SEEDS, function(SEED) {
    set.seed(SEED)
    mysample <- sample(1:37,n,replace=FALSE)*2
    mysample <- c(mysample,(mysample-1))
    mysample <- mysample[order(mysample)]
    design2 <- design[mysample,]
    design2 <- design2[,colSums(design2)>0]
    mval2 <- mval[,mysample]
    dm2 <- runlimma(mval=mval2,design=design2,myann=myann)
    dmagg2 <- agg(dm2,cores=2)
    m2 <- dmagg2$gme_res_df[,"mean",drop=FALSE]
    lamres2 <- runmitch(m=m2,genesets=gs_symbols,cores=2)
    lamsig2 <- subset(lamres2,p.adjustANOVA<0.05)
    lamsigsets2 <- paste(rownames(lamsig2),as.character(lamsig2$s.dist > 0))
    TOT=length(lamsigsets2)
    TP=length(which(lamsigsets2 %in% lamsigsets))
    FP=length(which(!lamsigsets2 %in% lamsigsets))
    FN=length(which(!lamsigsets %in% lamsigsets2))
    RES=c("TOT"=TOT,"TP"=TP,"FP"=FP,"FN"=FN)
    RES
  },mc.cores=5)
  do.call(rbind,dsres)
})

@@

dots <- lapply(ssres,function(x) { x[,"TP"] })
boxplot(dots,names=SAMPLESIZES,col="white",ylab="no. true positives",xlab="sample size",cex=0,main="LAM") ; grid()
beeswarm(dots,add=TRUE,pch=1,cex=0.6)
abline(h= ssres[[1]][1,1])

@@

head(dm, 10) %>% kbl(caption="Top 10 probes by p-value") %>% kable_paper("hover", full_width = F)

sigup <- rownames(subset(dm,logFC > 0 & adj.P.Val < 0.05))

sigdn <- rownames(subset(dm,logFC < 0 & adj.P.Val < 0.05))

@@

tic()
gsaup <- gsameth(sig.cpg=sigup, all.cpg=rownames(dm), collection=gs_entrez)
rownames(gsaup) <- paste(rownames(gsaup),"UP")
gsadn <- gsameth(sig.cpg=sigdn, all.cpg=rownames(dm), collection=gs_entrez)
rownames(gsadn) <- paste(rownames(gsadn),"DN")
toc() # 12.9, 12.6, 11.5

@@

gsig_up <-  subset(gsaup,FDR<0.05)
head(gsig_up,20) %>% kbl(caption="Hypermethylated pathways") %>% kable_paper("hover", full_width = F)

gsig_dn <-  subset(gsadn,FDR<0.05)
head(gsig_dn,20) %>% kbl(caption="Hypomethylated pathways") %>% kable_paper("hover", full_width = F)

gsig <- rbind(gsig_up,gsig_dn)

@@

gsres <- lapply(SAMPLESIZES,function(n) {
  SEEDS <- seq(100,5000,100)
  dsres <- mclapply(SEEDS, function(SEED) {
    set.seed(SEED)
    mysample <- sample(1:37,n,replace=FALSE)*2
    mysample <- c(mysample,(mysample-1))
    mysample <- mysample[order(mysample)]
    design2 <- design[mysample,]
    design2 <- design2[,colSums(design2)>0]
    #mval downsample
    mval2 <- mval[,mysample]
    dm2 <- runlimma(mval2,design2,myann)
    if ( is.na(dm2$logFC[1]) ) {
      RES=c("TOT"=0,"TP"=0,"FP"=0,"FN"=length(rownames(gsig)) )
    } else {
      sigup2 <- rownames(subset(dm2,adj.P.Val<0.05 & UCSC_RefGene_Name != "" & logFC>0))
      if ( length(sigup2) < 250 ) { sigup2 <- head(rownames(subset(dm2,logFC>0 & UCSC_RefGene_Name != "" )),250) }
      sigdn2 <- rownames(subset(dm2,adj.P.Val<0.05 & UCSC_RefGene_Name != "" & logFC<0))
      if ( length(sigdn2) < 250 ) { sigdn2 <- head(rownames(subset(dm2,logFC<0 & UCSC_RefGene_Name != "" )),250) }
      gsaup2 <- gsameth(sig.cpg=sigup2, all.cpg=rownames(dm2), collection=gs_entrez)
      rownames(gsaup2) <- paste(rownames(gsaup2),"UP")
      gsadn2 <- gsameth(sig.cpg=sigdn2, all.cpg=rownames(dm2), collection=gs_entrez)
      rownames(gsadn2) <- paste(rownames(gsadn2),"DN")
      gsig_up2 <- subset(gsaup2,FDR<0.05)
      gsig_dn2 <- subset(gsadn2,FDR<0.05)
      gsig2 <- rbind(gsig_up2,gsig_dn2)
      TOT=nrow(gsig2)
      TP=length(which(rownames(gsig2) %in% rownames(gsig)))
      FP=length(which(!rownames(gsig2) %in% rownames(gsig)))
      FN=length(which(!rownames(gsig) %in% rownames(gsig2)))
      RES=c("TOT"=TOT,"TP"=TP,"FP"=FP,"FN"=FN)
    }
    RES
  }, mc.cores=5)
  do.call(rbind,dsres)
})

mx <- matrix(rep(0,200),ncol=4)
colnames(mx) <- c("TOT","TP","FP","FN")

@@

blah()

### 2023-11-15 manuscript/manuscript.Rmd
progress on sensitivity analysis

The same dataset including all 37 patients underwent analysis using the LA method, gsameth and ebGSEA.
The effect size (enrichment score) of significantly differentially methylated pathways (FDR<0.05) was
recorded.
Then, a random subset of n patients was selected followed by pathway enrichment analysis with the
respective method.
The subset results were compared to the full n=37 results to test the sensitivity of these enrichment
methods at differenent effect size levels.
This process was repeated 20 times.

@@

We applied the three distinct approaches ("LA", "AL" and "AA") for pathway level differential
methylation analysis to a dataset consisting of lung tumour and adjacent normal tissue from 37 patients.
The dataset was split randomly into two groups of 18 patients and the overlap between subset was used to
assess the consistency and validity of findings with 50 simulations.
From the 1550 Reactome gene sets in the dataset, the AL method recorded the highest number of
"valid" pathways with an average of 919.8 being identified in both subsets, which was much higher than
LA and AA methods with 295.3 and 276.7 gene sets respectively (Figure 1A).
The AL method also recorded the lowest number of uncommon sets (382.0) which was better than the LA and
AA methods with 634.8 and 908.3 respectively (Figure 1B).
We also noticed that the LA method sometimes yielded contradictory results, so we quantified the number
of discordant observations, that is statistically significant but in the opposite direction.
True to this observation, the LA method gave 461.6 discordant pathways which indicates a high rate of
false-positives.
Meanwhile the AL and AA methods gave a mean of 0.1 and 0.02 discordant pathways respectively.
Taken together, these results indicate that the AL method gives the highest rate of common results
suggesting good accuracy in terms of sensitivity and specificity in this context.

>>>>

The same dataset including all 37 patients underwent analysis using the LA method, gsameth and ebGSEA.
The effect size (enrichment score) of significantly differentially methylated pathways (FDR<0.05) was
recorded.
Then, a random subset of n patients was selected followed by pathway enrichment analysis with the
respective method.
The subset results were compared to the full n=37 results to test the sensitivity of these enrichment
methods at differenent effect size levels.
This process was repeated 50 times.

@@

* LAM has better performance than GSAmeth

* Could be due to (1) filtering genes without annotations, and (2) FDR not applied for sets without members.

* Can make changes to the GSAmeth codes to fix the problems.
